---
output:
  pdf_document: default
  html_document: default
---

# Association rule mining in R

This lab activity is adapted from the following tutorials/guides:
- [Association mining (Market Basket Analysis)](http://r-statistics.co/Association-Mining-With-R.html)
- [Introduction to Association Rule Mining in R](https://www.kirenz.com/post/2020-05-14-r-association-rule-mining/)

In this lab activity, we will

- use the `arules` package to perform association rule mining
- load our transaction data from file (with the appropriate format for the `arules` package)

## Dependencies

We'll use the following packages in this lab activity (you will need to install any that you do not already have installed):

```{r}
library(tidyverse)  # For data wrangling
library(arules)     # Association rule mining algorithms/functions
library(arulesViz)  # Contains visualizations for association rule mining
```

## Data preparation and inspection

In this lab, we'll load our transaction data from file: `transactions.dat`, which you can download on blackboard or you can access it online [here](https://raw.githubusercontent.com/amlalejini/gvsu-cis635-2022f/main/lecture-material/week-11/transactions.dat).

Before continuing on with loading these data, take a look at `transactions.dat`.
Notice that each line (after the first) contains a comma separated list of items.
Each line describes a single transaction.
For example, the line `B,E,C` describes a transaction containing the items B, C, and E.

We can use the `read.transactions` function from the `arules` package to read a our data file containing transactions.

```{r}
# You will need to adjust the file path to run this lab activity locally.
transactions <- read.transactions(
  "lecture-material/week-11/transactions.dat",
  sep = ",",
  skip = 1
)
```

In your R console, run `?read.transactions` to see more information about loading transaction data from file.

The `arules` packages provides some useful functions for inspecting our transaction data:

To view the different items represented across all transactions in your data, you can use the `itemLabels` function:

```{r}
itemLabels(transactions)
```

To view the sizes of each transaction in our dataset, we can use the `size` function:

```{r}
# "arules::" before calling the size function tells R that we want to use the
# size function provided by the arules package.
arules::size(transactions)
```

If you wanted to get a list object containing all transactions, you could use the `LIST` function:

```{r}
transaction_list <- LIST(transactions)
# For brevity, we'll just show the first few entries in the list we created.
head(transaction_list)
```

The good old `summary` function works with transaction data, too:

```{r}
summary(transactions)
```

The `arules` package also provides the `image` function to visually inspect binary incidence matrices describing your data.
In most cases, you'll have too many transactions for this to be a particularly useful visualization.

```{r}
# look at all transactions
image(transactions)
```

The `arules` package also provides a quick function for visualizing the frequences of individual items in your transaction data:

```{r}
itemFrequencyPlot(transactions, topN=10, cex.names=1)
```

In these data, we can see that C is the most frequent and E is the least frequent.

## Generating association rules

The data provided in `transactions.dat` were generated randomly (for demonstrative purposes).
I.e., I wanted to show off a small example of transaction data formated in a way that works with the functions in the `arules` package.
Because these data were generated random, the association rules are not going to be particularly meaningful.

We can use the `apriori` function to generate association rules.
Run `?apriori` in your R consolue for more information about using and parameterizing the `apriori` function.

```{r}
rules <- apriori(
  transactions,
  parameter = list(
    supp = 0.3,     # Sets our minimum support threshold
    conf = 0.5,     # Sets our confidence threshold
    minlen = 2,     # Rules must have at least two items. Eliminates null rules.
    target = "rules"  # We'd like rules as our output.
  )
)
summary(rules)
```

We can use the `inspect` function to display the association rules that we found:

```{r}
inspect(rules)
```

### Generating frequent itemsets

What if you _just_ want to generate the set of frequent itemsets using the _apriori_ algorithm?
You can use the `apriori` function to do that if you change the target parameter:

```{r}
freq_itemsets <- apriori(
  transactions,
  parameter = list(
    supp = 0.3,     # Sets our minimum support threshold
    minlen = 1,     # Sets must have at least one item. Eliminates null sets.
    target = "frequent itemsets"
  )
)
summary(freq_itemsets)
```

Again, we can use the `inspect` function from the `arules` package to see all of the frequent itemsets:

```{r}
inspect(freq_itemsets)
```

## Exercises

- Adjust confidence and support thresholds used for rule generation. What happens if you decrease these thresholds? What about increase these threshold?
- Apply association rule mining to the Groceries dataset.
  The Groceries dataset comes with `arules` package. You can load it as follows:
```{r}
data("Groceries")
grocery_trans <- Groceries
```
- Describe a type of data that does not describe customer transactions that you could imagine applying association rule mining to.
  For example, how might you apply association rule mining to generate association rules for characters in a set of words?
  What other types of data might be useful to apply association rule mining to?


## References and further exploration

- [Introduction to Association Rule Mining in R by Jan Kirenz](https://www.kirenz.com/post/2020-05-14-r-association-rule-mining/)
- [`arules` package usage documentation](https://cran.r-project.org/web/packages/arules/vignettes/arules.pdf)
- [Association rule mining with R by Selva Prabhakaran](http://r-statistics.co/Association-M.ining-With-R.html)