---
output:
  pdf_document: default
  html_document: default
---

# Sequential pattern mining in R

The main goal of this lab activity is to provide a beginner-friendly introduction to performing sequence pattern mining in R using the `arulesSequences` package.
In this lab activity, we will

- use the `arulesSequences` package to perform sequential pattern mining
- load basic sequence data from file (with the appropriate format for the `arulesSequences` package)

## Dependencies

We'll use the following packages in this lab activity (you will need to install any that you don't already have installed):

```{r}
library(tidyverse)        # For data wrangling, visualization
library(arulesSequences)  # For sequential pattern mining algorithms
library(cowplot)          # For a nice ggplot theme
library(khroma)

theme_set(theme_cowplot()) # Set the cowplot ggplot theme as the default.
```

Find the documentation for the `arulesSequences` package here: <https://cran.r-project.org/web/packages/arulesSequences/index.html>.

## Data preparation and inspection

In this lab, we'll load our sequences from file: `sequences.dat`, which you can download on blackboard or you can access it online [here](https://raw.githubusercontent.com/amlalejini/gvsu-cis635-2022f/main/lecture-material/week-12/sequences.dat).

Before continuing on with loading these data, take a look at `sequences.dat`.
Each line describes _part_ of a sequence.
Notice that each line is formatted as follows:
sequence id (sequenceID), itemset id (eventID), itemset size (size), and then all items in the itemset separated by commas.

For example, then line `2,1,1,a,b` describes the first itemset in sequence 2; this itemset is size 2 with the following items: `a` and `b`.
Notice that you need to look at multiple lines to see all of the itemsets in sequence 2.

This isn't a well-formed .csv file (variable number of items trailing on each line), hence I have opted to use the `.dat` extension instead of a `.csv` extension.

We can use the `read_baskets` function from the `arulesSequences` package to read our file containing itemset sequences.

```{r}
# You may need to adjust the path to the sequences.dat file.
seqs <- arulesSequences::read_baskets(
  "lecture-material/week-12/sequences.dat",
  sep = ",",
  info = c("sequenceID", "eventID", "size")  # Identities of first N columns
)
```

We can use the `inspect` function provided in the `arulesSequences` package to view our data:

```{r}
arulesSequences::inspect(seqs)
```

If you want to view the set of items represented across all sequences in our data, you can use the `itemLabels` function:

```{r}
itemLabels(seqs)
```

To view the frequency of each item, you can use the `itemFrequency` function:

```{r}
itemFrequency(seqs)
```

## Finding frequent subsequences

The `arulesSequences` package provides an implementaiton of the cSPADE algorithm.
For more information about how the cSPADe algorithm works, see the following paper:
[Zaki. 2000. Sequence mining in categorical domains: incorporating constraints. In Proceedings of the ninth international conference on Information and knowledge management (CIKM '00). Association for Computing Machinery, New York, NY, USA, 422–429. https://doi.org/10.1145/354756.354849](https://doi.org/10.1145/354756.354849)

Here, we set our minimum support threshold to 0.75; that is, we require that subsequences appear in at least 75% of sequences in our dataset to be considered frequent.

```{r}
# Apply the cspade algorithm to our seqs dataset
freq_seqs <- cspade(
  seqs,
  parameter = list(
    support = 0.75
  )
)
```

The inspect function lets us view the frequent subsequences:

```{r}
inspect(freq_seqs)
```

We can convert the frequent subsequences into a data frame, which makes it easier to use tidyr and ggplot to further analyze the frequent subsequences.

```{r}
# Create a dataframe with the frequent subsequences we found
freq_seqs_df <- as(freq_seqs, "data.frame")
```

Using the dataframe version of our frequent subsequences, we can more easily use ggplot to visualize support for frequent subsequences in our dataset:

```{r}
# Adapted from: https://jlaw.netlify.app/2020/11/01/sequence-mining-my-browsing-history-with-arulessequences/
ggplot(
    freq_seqs_df,
    aes(
      x = fct_reorder(sequence, support),
      y = support,
      fill = sequence
    )
  ) +
  geom_col() +
  geom_label(
    aes(
      label = scales::percent(support)
    ),
    hjust = 0.5
  ) +
  labs(
    x = "Subsequence",
    y = "Support",
    title = "Frequent sequences",
    caption = "Support gives percent of sequences with given subsequence"
  ) +
  scale_fill_vibrant(
    guide = "none"
  ) +
  scale_y_continuous(
    labels = scales::percent,
    expand = expansion(mult = c(0, .1))
  ) +
  coord_flip()
```

### Rule generation

We can generate high-confidence rules from the frequent subsequences that we found using the `ruleInduction` function from the `arulesSequences` package.
Below, we compute all rules with confidence greater than 0.75.

```{r}
rules <- ruleInduction(
  freq_seqs,
  confidence=0.75
)
```

Again, we can use the `inspect` function to view the rules we found:

```{r}
inspect(rules)
```

## Exercises

- Adjust the support threshold used for finding frequent subsequences.
  What happens if you increase the support threshold? What happens if you decrease the support threshold?
- Describe a type of data that you think would be interesting to apply sequential pattern mining to.
  What do sequences represent? What does each itemset represent? What does each item represent? At a high level (not mathematically), what does it mean for a subsequence to be frequent in the context of your chosen domain?
- What would it mean to apply frequent subsequence mining to a database of words?
  You could assume each word is a sequence of size-1 itemsets.
- Apply sequential pattern mining to wordle data: [world.csv]([here](https://raw.githubusercontent.com/amlalejini/gvsu-cis635-2022f/main/lecture-material/week-12/wordle.csv))
  - To use the `cspade` function in `arulesSequences`, you will need to preprocess the data into the expected format (i.e., like `sequences.dat`).

## Further exploration

- In the following blog post, the author uses R to apply sequential pattern mining to their web browsing history: [Sequence Mining My Browsing History with arulesSequences](https://jlaw.netlify.app/2020/11/01/sequence-mining-my-browsing-history-with-arulessequences/)
- For more information on the cSPADE algorithm, see [Zaki. 2000. Sequence mining in categorical domains: incorporating constraints. In Proceedings of the ninth international conference on Information and knowledge management (CIKM '00). Association for Computing Machinery, New York, NY, USA, 422–429. https://doi.org/10.1145/354756.354849](https://doi.org/10.1145/354756.354849)

## References

- `arulesSequences` Documentation: <https://cran.r-project.org/web/packages/arulesSequences/index.html>
- [Sequence Mining My Browsing History with arulesSequences](https://jlaw.netlify.app/2020/11/01/sequence-mining-my-browsing-history-with-arulessequences/) by [JLaw](https://jlaw.netlify.app/)
- Zaki. 2000. Sequence mining in categorical domains: incorporating constraints. In Proceedings of the ninth international conference on Information and knowledge management (CIKM '00). Association for Computing Machinery, New York, NY, USA, 422–429. https://doi.org/10.1145/354756.354849